Think deep for reef resilience

### 2025-08-14 paper/paper_draft.qmd
Update remaining commented paragraphs and figure caption for depths

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding which reefs exhibit resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover timeseries and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintainence of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-timeseries resilience and
    maintainence of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
editor:
  markdown:
    wrap: 92
---

@@

The results presented herein suggest that when modelled as part of a large and complex
ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide
evidence in support of depth as a form of thermal refugia for shallow-water reefs on the
GBR. While protecting ‘key node’ reefs in larval connectivity networks is an important
consideration for reef restoration efforts [@anthony2020; @hock2017], our results suggest
that current connectivity modelling methods are not able to generate connectivity that
contributes to reef resilience under projected conditions **\[TI: TO BE REVISITED\]**. We
suggest further analyses that investigate the benefits of incorporating depth-based
resilience into reef decision making, as well as further research and development in
connectivity modelling methods for whole-ecosystem scales.

@@

![Depth cutoff for
reefs](../outputs/ADRIA_results/HighResCoralStress/figs/reef_depth_included_cutoff.png){#sfig-depth-cutoff
fig-align="center" width="14.8cm" height="9.5cm"}

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover timeseries and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintainence of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-timeseries resilience and
    maintainence of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
editor:
  markdown:
    wrap: 92
---

@@

The results presented suggest that when modelled as part of a large and complex ecosystem,
reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in
support of depth as a form of thermal refugia for shallow-water reefs on the GBR. While
protecting ‘key node’ reefs in larval connectivity networks is an important consideration
for reef restoration efforts [@anthony2020; @hock2017], current connectivity modelling
methods, and our modelling framework, suggest that connectivity may not be as important for
reef resilience under projected conditions as factors such as depth **\[TI: TO BE
REVISITED\]**. We suggest further analyses that investigate the benefits of incorporating
depth-based resilience into reef intervention decision making, as well as further research
and development in connectivity modelling methods for whole-ecosystem scales.

@@

![Distribution of depth data coverage levels across 3806 possible reefs. Reefs below the 5%
data coverage threshold (represented by the red vertical line) were removed from the
investigation.](../outputs/ADRIA_results/HighResCoralStress/figs/reef_depth_included_cutoff.png){#sfig-depth-cutoff
fig-align="center" width="14.8cm" height="9.5cm"}
